Combinatorial interpretations of holonomic recurrences

### 2021-08-21 article/main.tex
Resolve multiple todos

\RequirePackage{geometry}
\RequirePackage{tabularx}
\RequirePackage{graphicx}
\RequirePackage{svg}

@@

\RequirePackage{amsthm}
\RequirePackage{amsmath}
\RequirePackage{amsfonts}
\RequirePackage{soul}
\RequirePackage{xspace}
\RequirePackage{bussproofs}
\RequirePackage[customcolors]{hf-tikz}

@@

Green edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{green_edge}
\end{minipage}
depict edges to \(\no\) nodes.

@@

\begin{figure}[h]
    \begin{center}
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B\)\\
                \includesvg[scale=0.5]{binary_tree_def_1}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(=\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(\text{leaf}\)\\
                \includesvg[scale=0.5]{binary_tree_def_2}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(+\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B^2\)\\
                \includesvg[scale=0.5]{binary_tree_def_3}%
            \end{center}
        \end{minipage}%
    \end{center}
    \caption{Binary tree recursive definition}
    \label{fig:binary_recursion}
\end{figure}

@@

\begin{center}
    \includesvg[scale=.7]{iilll}
\end{center}
\end{example}

@@

\begin{definition}
    Pre-order notation of \(n\)-ary tree \(T\), \(\enc{T}\), is defined recursively as \(\enc{\troot(T)} \prod_{i=1}^{n} \enc{i\text{-th subtree of } \troot(T)}\), ie.\ catenation of identifier of the root of the tree with encoded its subtrees from left to right.
\end{definition}

\begin{figure}[h]
    \centering
    \includesvg[scale=.7]{oononnn}
    \caption{\(\n \n \no \n \no \no \no\)}
    \label{fig:oononnn}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{remy_left}\end{minipage}%
    \caption{\(\enc{R_l(T, v)}\)}
    \label{fig:remy_left}
\end{figure}

\begin{figure}[h]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{remy_right}\end{minipage}%
    \caption{\(\enc{R_r(T, v)}\)}
    \label{fig:remy_right}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includeinlinesvg{.2}{unary_binary_base}%
    \(\Rightarrow\)%
    \includeinlinesvg{.3}{unary_binary_full}
    \caption{Unary-binary tree to full-binary tree transformation}
    \label{fig:unary_binary_transformation}
\end{figure}

@@

\begin{center}
    \begin{tabular}{| c | c | c |}
        \hline

@@

        &
        I \includesvg[scale=0.4]{ub_1si} &
        Vro \includesvg[scale=0.4]{ub_0fid} \\
        \hline
    \end{tabular}
\end{center}
\todo{finish the table}

@@

\begin{figure}[h]
    \centering
    \includesvg[scale=0.4]{schroder_1}
    \includesvg[scale=0.4]{schroder_2}
    \includesvg[scale=0.4]{schroder_3}
    \includesvg[scale=0.4]{schroder_4}
    \(\ldots\)
    \caption{Schröder tree node}
    \label{fig:schroder_trees}
\end{figure}

@@

\todo{image of the transformations}

@@

\todo{image of the problematic cases}

@@

\todo{images}

In case 1. and 2. we can just swap the labels as follows:

\todo{images}

@@

To achieve trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3. and replace the father of the \(s\) with \(t''\) and point root of \(t''\). In this way, we will loose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\).

\todo{image}

@@

\todo{describe binary trees example}

The method itself has proven to be helpful in order to divide the problem into smaller parts. Unfortunately It does not give any immediate interpretation for applying the inductive step.

@@

\includeinlinesvg{.19}{trees_00__4_base}%
\(\Rightarrow\)
\includeinlinesvg{.19}{trees_00__4}%
\includeinlinesvg{.19}{trees_00__5}%
\includeinlinesvg{.19}{trees_00__6}%
\includeinlinesvg{.19}{trees_00__7}%

@@

\includeinlinesvg{.33}{trees_00__2}%
\includeinlinesvg{.33}{trees_00__3}%
\includeinlinesvg{.33}{trees_00__7}%

\includeinlinesvg{.33}{trees_00__4}%
\includeinlinesvg{.33}{trees_00__12}%
\includeinlinesvg{.33}{trees_00__13}%

\includeinlinesvg{.16}{trees_00__0}%
\includeinlinesvg{.16}{trees_00__1}%
\includeinlinesvg{.16}{trees_00__5}%
\includeinlinesvg{.16}{trees_00__6}%
\includeinlinesvg{.16}{trees_00__14}%
\includeinlinesvg{.16}{trees_00__15}%

\includeinlinesvg{.25}{trees_00__8}%
\includeinlinesvg{.25}{trees_00__9}%
\includeinlinesvg{.25}{trees_00__10}%
\includeinlinesvg{.25}{trees_00__11}%

@@

\includeinlinesvg{.24}{trees_100__0_base}\(\Rightarrow\)\includeinlinesvg{.24}{trees_100__0}%
\includeinlinesvg{.24}{trees_100__1_base}\(\Rightarrow\)\includeinlinesvg{.24}{trees_100__1}%

\includeinlinesvg{.24}{trees_100__2_base}\(\Rightarrow\)\includeinlinesvg{.24}{trees_100__2}%
\includeinlinesvg{.24}{trees_100__3_base}\(\Rightarrow\)\includeinlinesvg{.24}{trees_100__3}%

@@

\includeinlinesvg{.32}{trees_111__1_base}%
\(\Rightarrow\)
\includeinlinesvg{.32}{trees_101__0}%
\includeinlinesvg{.32}{trees_112__0}%

@@

\includeinlinesvg{.20}{trees_200__0_base}%
\(\Rightarrow\)
\includeinlinesvg{.20}{trees_200__0}%
\includeinlinesvg{.20}{trees_201__0}%
\includeinlinesvg{.20}{trees_202__0}%
\includeinlinesvg{.20}{trees_203__0}%

>>>>

\RequirePackage{geometry}
\RequirePackage{tabularx}
\RequirePackage{longtable}
\RequirePackage{graphicx}
\RequirePackage{svg}

@@

\RequirePackage{amsthm}
\RequirePackage{amsmath}
\RequirePackage{mathtools}
\RequirePackage{amsfonts}
\RequirePackage{soul}
\RequirePackage{xspace}
\RequirePackage{bussproofs}
\RequirePackage[customcolors]{hf-tikz}

@@

Dashed edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{intro__dashed_edge}
\end{minipage}
depict edges that may or may not exist (eg.\ when we describe transformations and we don't care if the node is root or if it has children).

Green edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{intro__green_edge}
\end{minipage}
depict edges to \(\no\) nodes.

@@

\begin{figure}[h]
    \begin{center}
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B\)\\
                \includesvg[scale=0.5]{binary__def_1}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(=\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(\text{leaf}\)\\
                \includesvg[scale=0.5]{binary__def_2}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(+\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B^2\)\\
                \includesvg[scale=0.5]{binary__def_3}%
            \end{center}
        \end{minipage}%
    \end{center}
    \caption{Binary tree recursive definition}
    \label{fig:binary_recursion}
\end{figure}

@@

\begin{center}
    \includesvg[scale=.7]{intro__iilll}
\end{center}
\end{example}

@@

\begin{definition}
    Pre-order notation of \(n\)-ary tree \(T\), \(\enc{T}\), is defined recursively as:
    \[\enc{T} = \enc{\troot(T)} \prod_{i=1}^{n} \enc{i\text{-th subtree of } \troot(T)}\]
    ie.\ catenation of identifier of the root of the tree with encoded its subtrees from left to right.
\end{definition}

\begin{figure}[h]
    \centering
    \includesvg[scale=.7]{intro__oononnn}
    \caption{\(\n \n \no \n \no \no \no\)}
    \label{fig:oononnn}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_left}\end{minipage}%
    \caption{\(\enc{R_l(T, v)}\)}
    \label{fig:remy_left}
\end{figure}

\begin{figure}[h]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_right}\end{minipage}%
    \caption{\(\enc{R_r(T, v)}\)}
    \label{fig:remy_right}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includeinlinesvg{.2}{unary_binary__base}%
    \(\Rightarrow\)%
    \includeinlinesvg{.3}{unary_binary__full}
    \caption{Unary-binary tree to full-binary tree transformation}
    \label{fig:unary_binary_transformation}
\end{figure}

@@

\begin{center}
    \begin{longtable}{| c | c | c |}
        \hline

@@

        &
        I \includesvg[scale=0.4]{unary_binary__1si} &
        Vro \includesvg[scale=0.4]{unary_binary__0fid} \\
        \hline

        &
        Vlt \includesvg[scale=0.4]{unary_binary__1fjg} &
        VrtUrv \includesvg[scale=0.4]{unary_binary__0fjdOgf} \\
        \hline

        I \includesvg[scale=0.4]{unary_binary__2i} &
        &
        VrtUra \includesvg[scale=0.4]{unary_binary__0fjdOdi} \\
        \hline

        &
        Vro \includesvg[scale=0.4]{unary_binary__1fid} &
        VrtUla \includesvg[scale=0.4]{unary_binary__0fjdOgi} \\
        \hline

        &
        Vrt \includesvg[scale=0.4]{unary_binary__1fjd} &
        VltUrv \includesvg[scale=0.4]{unary_binary__0fjgOdf} \\
        \hline

        I \includesvg[scale=0.4]{unary_binary__2i} &
        &
        VltUra \includesvg[scale=0.4]{unary_binary__0fjgOdi} \\
        \hline

        I \includesvg[scale=0.4]{unary_binary__2i} &
        &
        VltUla \includesvg[scale=0.4]{unary_binary__0fjgOgi} \\
        \hline
    \end{longtable}
\end{center}

@@

\begin{figure}[h]
    \centering
    \includesvg[scale=0.4]{schroder__example_1}
    \includesvg[scale=0.4]{schroder__example_2}
    \includesvg[scale=0.4]{schroder__example_3}
    \includesvg[scale=0.4]{schroder__example_4}
    \(\ldots\)
    \caption{Schröder tree node}
    \label{fig:schroder_trees}
\end{figure}

@@

\begin{figure}[h]
    \begin{center}
    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{L_1}\)%
    \includeinlinesvg{.16}{schroder__l1}%
    \hspace{.1\textwidth}%
    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{L_2}\)%
    \includeinlinesvg{.16}{schroder__l2}%

    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{R_1}\)%
    \includeinlinesvg{.16}{schroder__r1}%
    \end{center}
    \caption{Transformations \(L_1\), \(L_2\) and \(R_1\)}
    \label{fig:transformations_L1_L2_R1}
\end{figure}

@@

\begin{center}
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.5]{schroder__case_a}\\
            Case 1.
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.5]{schroder__case_b}\\
            Case 2.
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.5]{schroder__case_c}\\
            Case 3.
        \end{center}
    \end{minipage}%
\end{center}

@@

\begin{center}
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.45]{schroder__case_a_l}\\
            Case 1.
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.45]{schroder__case_b_l}\\
            Case 2.
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.45]{schroder__case_c_l}\\
            Case 3.
        \end{center}
    \end{minipage}%
\end{center}

In case 1.\ and 2.\ we can just swap the labels as follows:

\begin{center}
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.45]{schroder__case_a_l_fixed}\\
            Case 1.
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.45]{schroder__case_b_l_fixed}\\
            Case 2.
        \end{center}
    \end{minipage}%
\end{center}

@@

To achieve trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3.\ and replace whole subtree with subtree of \(t''\) with its root pointed. In this way, we will loose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\):

\begin{center}
    \begin{minipage}[t]{.3\textwidth}
        \begin{center}
            \includesvg[scale=0.45]{schroder__case_c_l_fixed}\\
            Case 3.
        \end{center}
    \end{minipage}%
\end{center}

@@

For case of binary trees with size defined as number of leaves, we can define 2 recurrences:

\[\text{Holonomic} := nB(n) = 2 (2n - 3)B(n - 1)\]
\[\text{Non-linear} := B(n) = \sum_{i=1}^{n - 1} B(i) B(n - i)\]

The non-linear one has very simple interpretation. We can take 2 smaller trees and join them using newly created common root.

We can multiply both sides of Non-linear recurrence by \(n\) and break up \(n = i + (n - i)\) to get two similar sigmas:

\[n B(n) = \sum_{i=1}^{n - 1} i B(i) B(n - i) + \sum_{i=1}^{n - 1} (n - i) B(n - i) B(i)\]

Using our inductive hypothesis, ie.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such simple case as binay trees is almost trivial, especially if one has seen the solution once.

The method itself has proven to be helpful in order to divide the problem into smaller parts and find some corner cases. Unfortunately, it does not give any immediate interpretation for applying the inductive step.

@@

\includeinlinesvg{.19}{lambda__trees_00__4_base}%
\(\Rightarrow\)
\includeinlinesvg{.19}{lambda__trees_00__4}%
\includeinlinesvg{.19}{lambda__trees_00__5}%
\includeinlinesvg{.19}{lambda__trees_00__6}%
\includeinlinesvg{.19}{lambda__trees_00__7}%

@@

\includeinlinesvg{.33}{lambda__trees_00__2}%
\includeinlinesvg{.33}{lambda__trees_00__3}%
\includeinlinesvg{.33}{lambda__trees_00__7}%

\includeinlinesvg{.33}{lambda__trees_00__4}%
\includeinlinesvg{.33}{lambda__trees_00__12}%
\includeinlinesvg{.33}{lambda__trees_00__13}%

\includeinlinesvg{.16}{lambda__trees_00__0}%
\includeinlinesvg{.16}{lambda__trees_00__1}%
\includeinlinesvg{.16}{lambda__trees_00__5}%
\includeinlinesvg{.16}{lambda__trees_00__6}%
\includeinlinesvg{.16}{lambda__trees_00__14}%
\includeinlinesvg{.16}{lambda__trees_00__15}%

\includeinlinesvg{.25}{lambda__trees_00__8}%
\includeinlinesvg{.25}{lambda__trees_00__9}%
\includeinlinesvg{.25}{lambda__trees_00__10}%
\includeinlinesvg{.25}{lambda__trees_00__11}%

@@

\includeinlinesvg{.24}{lambda__trees_100__0_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__0}%
\includeinlinesvg{.24}{lambda__trees_100__1_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__1}%

\includeinlinesvg{.24}{lambda__trees_100__2_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__2}%
\includeinlinesvg{.24}{lambda__trees_100__3_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__3}%

@@

\includeinlinesvg{.32}{lambda__trees_111__1_base}%
\(\Rightarrow\)
\includeinlinesvg{.32}{lambda__trees_101__0}%
\includeinlinesvg{.32}{lambda__trees_112__0}%

@@

\includeinlinesvg{.20}{lambda__trees_200__0_base}%
\(\Rightarrow\)
\includeinlinesvg{.20}{lambda__trees_200__0}%
\includeinlinesvg{.20}{lambda__trees_201__0}%
\includeinlinesvg{.20}{lambda__trees_202__0}%
\includeinlinesvg{.20}{lambda__trees_203__0}%
